Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-09 aistats.tex
Updates from Overleaf

\usepackage[capitalise]{cleveref}
\newcommand{\RLP}[1]{\textcolor{red}{RLP:#1}}
\newcommand{\fdk}[1]{\textcolor{Periwinkle}{fdk:#1}}
\newcommand{\TODO}[1]{\textcolor{cyan}{TODO #1}}

% my packages
\usepackage{math_commands}
% some custom math commands
\newtheorem{proposition}{Proposition}
\newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
\usepackage{suffix}
\WithSuffix\newcommand\expect*[2][]{\ensuremath{\mathbb{E}_{#1} [ #2 ] }} % expectation operator

\newcommand{\cond}{\,\vert\,}
\newcommand{\logpart}{A}
\newcommand{\conj}{\logpart^*}
\newcommand{\bregman}{\cB_\logpart}
\newcommand{\bregmanconj}{\cB_{\logpart^*}}
\newcommand{\nat}{\theta}
\newcommand{\m}{m}
\newcommand{\meanp}{\m}
\newcommand{\decrement}{D}
\newcommand{\linear}{\ell} % linearization of a function
\newcommand{\lr}{\gamma} % learning rate, or step-size
\newcommand{\lin}[1]{\left\langle#1\right\rangle}

\newcommand{\MAPm}{\hat \m_n}
\newcommand{\MAPt}{\hat \nat_n}
\DeclareMathSymbol{\shortminus}{\mathbin}{AMSa}{"39}

\hypersetup{draft}

@@

\aistatstitle{Convergence Rates for the MAP of an Exponential Family, and Stochastic Mirror Descent -- an Open Problem}

@@

\begin{abstract}
We consider the problem of upper bounding the expected sub-optimality of the maximum likelihood estimate (MLE), or a conjugate maximum a posteriori (MAP) for an exponential family. 
Surprisingly, we found no general solution to this problem in the literature. 
In particular, current theories do not hold for a Gaussian, or in the very interesting few samples regime.
After exhibiting various facets of the problem, we show the MAP can be interpreted as the result of Stochastic Mirror Descent (SMD), 
yet it falls out of scope for modern analysis.
We believe solving this very fundamental problem may bring progress to both the statistics and optimization communities.
\end{abstract}

%In particular, no rates hold in the few samples regime  -- e.g. after seeing 5 samples from a gaussian, we do not know how many bits away from the true distribution we should expect our model to be.
%No rates hold either when the loss is ill-behaved : infinite domain, not smooth, not self-concordant

@@

But our understanding of stochastic optimization lacks in other domains. 
A generalization of the above problem, at the intersection of optimization and statistics, 
is the problem of minimizing 
\begin{equation}
	f(\theta) = \expect[x \sim p(x\cond\theta_*)]{-\log p(x\cond\theta)},
\end{equation}
When $p(x\cond\theta)$ is an exponential family distribution,
$-\log p(x\cond\theta) = A(\theta) - \lin{t(x), \theta}$ (omitting the reference distribution).
Here, $A$ is the log-partition function, a Legendre function that we assume can be minimized in closed form.
The Gaussian example is the special case $A(\theta) = \frac{1}{2}\norm{\theta}^2$.

There are algorithms for this setting, for example the MLE -- the (online) maximum likelihood after seeing $t$ samples -- or adding regularization, the MAP -- the maximum a posteriori.
As we show, the MAP can be interpreted as online mirror descent with decreasing step-sizes.
Despite recent advances in this field, our convergence rates are still lacking; they are either asymptotic, cover only special cases or require having seen many ($n \gg d$) samples.

@@

We are aware that \emph{a} solution to this problem would be the maximum likelihood estimate for which there is a wealth of asymptotic and large sample results \citep{van2000asymptotic,ostrovskii2021finite}. %TODO complete 
However, none of the theory we found could address for instance the case of a Gaussian with unknown mean and variance $\cN(\mu,\sigma^2)$, hence our contributions. 
%TODO kind of redundant "problem - question - problem" paragraphs, not disentangling stats and optim. To improve

\paragraph{Contributions}
\begin{itemize}
	\item We provide insights on different facets of this problem : quadratic approximations ; bias variance decomposition; and a detailed analysis of the Gaussian variance estimation problem.
	\item We show the connections between the optimization and statistical estimation view of the problem and discuss where current theories fail to fully capture the problem.
\end{itemize}

@@

Exponential Families are to statistics what linear models are to regression.
Many classical model are exponential families, and their convexity and modeling flexibility make them one of the most widely used statistical models. 
Here we review some of their properties. 
We point the reader towards \citet[Chapter 3]{wainwright2008graphical} for a full review.
The exponential family for data $X\in \cX$ with natural (or primal) parameter $\nat$  is specified by a sufficient statistic $T: \cX \rightarrow \real^d$ and a density with respect to a base measure $\nu(dx)$
\begin{equation}
	 p(X|\nat) = \exp( \nat^\top T(X) - \logpart(\nat)) \; ,
\end{equation}
Here $\logpart$ is the log-partition function -- i.e. the normalization term
\begin{align}
    \logpart(\nat) = \log \int e^{\nat^\top T(x)} \nu(dx) \; .
\end{align}
When explicit, we will write the random variable $T = T(X)$ for lightness.
For convenience, we focus on steep exponential families with $\logpart$ a convex function of Legendre type.
In particular, the set $\nat \in \Theta$ such that $\logpart(\nat) < \infty$ is open and convex.
This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}, \nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.

% TODO sort out relationships between : steep family, minimal statistics and Legendre type. What implies what ?

\paragraph{Duality}
The logpartition function $\logpart$ verifies the two following identities
\begin{align}
    \nabla\logpart(\nat) &=  \expect[p(T|\nat)]{T} =: \meanp \\
    \nabla^2 \logpart(\nat) &= \Cov_\nat[T] \geq 0
\end{align}
where $\meanp$ is called the mean (or dual) parameter.
From now on, we assume that the sufficient statistic $T$ is minimal 
-- e.g. its covariance is definite positive for all $\nat$ -- 
then the log-partition function $\logpart$ is strictly convex and its gradient $\nabla \logpart$ is a bijection between natural parameters $\nat$ and mean parameters $\m$.
$\meanp$ lives in the open convex set $\cM$ equal to the relative interior of the convex hull of $T(\cX)$.
We will write interchangeably $\m$ or  $\nat$ depending on the context, being aware that both represent the same distribution.
At this point it is useful to introduce the convex conjugate (a.k.a. Fenchel-Legendre transform) of the logpartition function
\begin{align}
	\conj(\m) = \langle \m, \nat \rangle - \logpart(\nat) \; .
\end{align}
which matches the common notion of \textit{entropy} in information theory.
The entropy is eating mean parameters, and its gradient is mapping back to natural parameters $\nabla\conj \circ \nabla\logpart(\nat) = \nat$ (cf Fig.~\ref{fig:duality}).

\begin{figure}[ht]
	\centering
	\includegraphics{duality}
	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between natural and mean parameters $\nat, \meanp$. Figure copied from \citet{kunstner2020homeomorphic}. % TODO make an original one.
	}
	\label{fig:duality}
\end{figure}

\paragraph{Bregman Divergences}
A very useful object to keep in mind is the Bregman divergence induced by $\logpart$ between two parameters $\nat$ and $\nat_0$
\begin{align}
    \bregman (\nat ; \nat_0)
    & = \logpart(\nat) - \logpart(\nat_0) 
    - \langle \nabla \logpart(\nat_0)  , \nat - \nat_0 \rangle
\end{align}
with $\nabla \logpart(\nat_0) = \expect[\nat_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\nat_0$. 
It is a non-symmetric measure of distance between parameters.
It is equal to the divergence of its convex conjugate with switched arguments
\begin{align}
	\bregman (\nat ; \nat_0)
    = \bregmanconj ( \meanp_0 ; \meanp) \; .
\end{align}

\paragraph{Conjugate Prior}
One conjugate prior \citep{agarwal2010geometric} for $p(X|\nat)$ is
\begin{align}
    p(\nat) 
    &\propto \exp( - n_0 \bregman(\nat ; \nat_0) ) \\
    &\propto \exp(n_0 \m_0^\top \nat - n_0 \logpart(\nat))
\end{align}
where $n_0$ and $\nat_0$ are (hyper)parameters of the prior.
This is the formula for the exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and with natural parameter $(n_0 \m_0, -n_0)$.
Intuitively, $n_0$ is a number of fictive data points observed from a distribution with parameter $\nat_0$.

\paragraph{Maximum A Posteriori (MAP).}
Using Bayes rule, the posterior of $\nat$ given a dataset $\mD_n =(X_1,\dots,X_n)$ is
\begin{align}
	p(\nat \cond \mD_n)
	%\propto p(\mD|\nat)p(\nat) 
    \propto \exp(- (n_0+n) \bregman(\nat; \MAPt))
    \label{eq:joint_likelihood}
\end{align}
reaching its maximum in $\MAPt$, the MAP estimate verifying the first order optimality conditions
\begin{align}
    \nabla \logpart(\MAPt) = \MAPm
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
\end{align}
where $T_i=T(X_i)$.
When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)
\begin{align}
	\hat \m_n^\text{MLE} = \frac{\sum_{i=1}^n T_i}{n}
\end{align}
This last equation is known as moment matching.
The MLE and MAP estimates are statistics of the dataset $\mD_n$. 
Given a random dataset, we wish to bound their deviation from the optimum $\nat^*$ or $\meanp^*$.

@@

Let $\nat^* := \argmin \expect[X\sim\cD]{-\log p(X|\nat)}$, where $\cD$ is any distribution such that $\E[T] = \meanp^* \in \cM$.
Then the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
\begin{multline}
    \expect[X\sim \cD]{-\log p(X|\nat) + \log p(X|\nat^*) } \\
	= \KL( p(.|\nat^*) ; p(.|\nat)) \; .
\end{multline}
Note that $\cD$ does not have to be part of the exponential family.
For the exponential family, the KL is equal to the Bregman divergence induced by the log-partition function, or alternatively by the entropy, being careful with the order of the arguments 
\begin{align}
\boxed{
	\KL( p(.|\nat^*) ; p(.|\nat))
    = \bregman (\nat ; \nat^*)
    = \bregmanconj ( \meanp^* ; \meanp)
}
\end{align}

@@

Our question is: how does the KL behave when $\nat$ is the maximum-likelihood or the MAP estimate ? Can we get upper-bounds on the following quantities ?
\begin{align}
	\label{eq:bregmanMLE}
	\expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )} \\
	\label{eq:bregmanMAP}
	\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )} 
\end{align}
where the outer expectation is on the dataset $\mD = \{X_1, \dots, X_n \}$ sampled i.i.d from $\cD$.
More explicitly, can we write an upper bound that does not involve an expectation over the dataset -- e.g. where the stochasticity is removed.

\paragraph{Remark.}
This could be seen as a concentration inequality, expressed with a Bregman divergence instead of a norm.
However there is a connection between the random variable $T(X)$ and the metric $\logpart$. 
Indeed expressions~\eqref{eq:bregmanMLE} or~\eqref{eq:bregmanMAP} can be infinite for another choice of random variable. 
For instance, if we plug in $\conj(\m)= -\log(\m)$, which defines a divergence on positive numbers, and $T(X) \sim \cN(0,1)$ which can be negative.

@@

\subsection{MAP as Stochastic Mirror Descent}
\label{ssec:MAP=SMD}
The posterior~\eqref{eq:joint_likelihood} can always be written iteratively $p(\nat | \mD_n) \propto p(X_n | \nat) p(\nat | \mD_{n-1})$, so that the MAP verifies
\begin{align*}
	\hat \nat_{n}
	= \argmin_\nat - \log p(X_{n} | \nat) + (n_0 + n -1) \bregman(\nat; \hat \nat_{n-1})
    %\label{eq:SBPP}
\end{align*}
which is the update of a Stochastic Bregman Proximal Point update with step-size $\inv{n_0 + n - 1}$. Expanding $p(X_{n} | \nat)$ and  playing with the Bregman we get
\begin{align}
		\hat \nat_{n}
	= \argmin_\nat g_{n}^\top \nat + (n_0 + n) \bregman(\nat; \hat \nat_{n-1})
    \label{eq:SMD}
\end{align}
where $g_{n} := \nabla\logpart(\hat \nat_{n-1}) - T(X_n) = \hat \m_{n-1} - T_n$ is the gradient of $-\log p(X_{n} | \nat)$ evaluated at $\hat \nat_{n-1}$. 
We recognize the update of Stochastic Mirror Descent (a.k.a. Stochastic Bregman Gradient) with step-size $\lr_n = \inv{n_0 + n}$ -- a stochastic gradient step in the dual $\hat \m_n = \hat \m_{n-1} - \lr_n g_n$.
This parallel means it would be possible to obtain a finite convergence rate for the  MAP from an analysis of these algorithms. 
We give a special focus on this approach in Section~\ref{sec:optimization}.

\subsection{Asymptote}
As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
Asymptotic results typically show that natural parameters $\nat$ asymptotically normal ; reaching the Cramer-Rao lower bound \citep[for instance Ch4.2]{van2000asymptotic}.
Here the MAP is more simply expressed with mean parameters $\meanp$.
We get the asymptote by approximating the Bregman divergence using a second order Taylor expansion
\begin{align}
    \bregmanconj(\m^* ; \m) 
    &=\half  \norm{\m^* - \m}^2_{\nabla^2\conj(\m^*)}
    + O(\norm{\m - \m^*}^3)
\end{align}
where the Mahalanobis norm  $\| x \|_\mM^2 = x^\top \mM x$  is induced by the hessian of the entropy -- the inverse \textit{Fisher information matrix} -- at the optimum, 
\begin{align}
    \mF^*
    :=\nabla^2\conj(\m^*) 
    = \nabla^2\logpart(\nat^*)^{-1} 
    = \Cov_{\nat^*}[T(X)]^{-1}  \; .
\end{align}
From there, it is possible to show an asymptotic rate for the MLE and the MAP
\begin{align}
\label{eq:MAP_asymptote}
	\E \bregmanconj \left (\E [T(X)] ; \hat \meanp_n^\text{MLE/MAP} \right ) 
	= \frac{d}{2n} + O(n^{- \frac{3}{2}}) \; .
\end{align}
Both MLE and MAP have the same asymptote, as the contribution of the prior $n_0 \meanp_0$ gets negligible compared to the data for large $n$.
This asymptote is also independent of the optimum $\meanp^*$.

\subsection{Quadratic Case}
As another reference point, let us consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
For instance, this is the log-partition of a Gaussian with variance $1$ and unknown mean ($\cX=\real, \nu(dx) = \exp(\half[-x^2])dx, T(X)=X$).
Then $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances
\begin{align}
	\bregmanconj(\meanp^* ; \meanp) = \half \norm{\meanp^* -  \meanp }_2^2  \; .
\end{align}
Thanks to the independence of samples we can break down the MLE into individual point's contributions
\begin{align}
	\expect{\half \norm{\m^* -  \inv{n}  \smallsum_i T_i}_2^2} 
	=\frac{\Var(T)}{2n} 
\end{align}
Adding a reference mean $\m_0$ to get the MAP yields
\begin{align}
	\expect{\bregmanconj(\meanp^*; \MAPm)}
	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{(n+n_0)^2}
	%\\ &= O\left(\frac{\Var(T)}{n} \right) + O\left(\frac{\norm{\m^* -  \m_0}^2}{n^2} \right)
	\label{eq:MAP_quadratic}
\end{align}
so we have a variance term in $O(n^{-1})$ and a bias term decreasing as $O(n^{-2})$. We would like to get a similar result for arbitrary exponential families.
If we make restrictive assumptions on the log-partition function, we can relate $\bregmanconj$ to a norm or a quadratic.

{\bf If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
\begin{align}
    \bregmanconj(\m^* ; \m) 
    &\leq L \norm{\m^* - \m} + \norm{\nat} \norm{\m^* - \m} \\
    &\leq 2L \norm{\m^* - \m}
\end{align}
so $\bregmanconj$ is $2L$-Lipschitz, and we get a $O(\inv{\sqrt{n}})$ rate.

{\bf If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $\frac{1}{L}$-strongly convex), then
\begin{align}
    \bregmanconj(\m^* ; \m) 
    \leq \frac{L}{2} \norm{\m^* - \m}^2
\end{align}
so $\bregmanconj$ is upper bounded by a quadratic, and we get~\eqref{eq:MAP_quadratic} as an upper bound.

\subsection{Locally Quadratic Case}
All Bregman divergences are locally quadratic. 
Under some assumptions, we can quantify  when this quadratic behavior  kicks in.
For instance {\bf when $\conj$ is self-concordant} \citep[Ch.4.1]{nesterov2003introductory}. 
This is true when $T$ is 1-dimensional and $\logpart$ is self-concordant 
-- e.g. exponential distribution, 
Gaussian\fdk{with known variance?} and Laplace distributions with known mean,
Pareto distribution with known minimum value, 
Weibull distribution with known shape $k$
-- or when $T$ lives in a compact \citep{bubeck2015entropic}. Then the bound
\begin{align}
	 \bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF^*}^2
\end{align}
holds whenever $\norm{\m^*-\m}_{\mF^*} < 0.21$.\fdk{where does this $0.21$ come from?}
Similarly, \citet{kakade2010learning} assumes a bound on all higher order moments at $\nat^*$ to quantify when $\logpart$ starts behaving quadratically, but their rate does not directly apply to $\conj$ and $\m$.

\TODO{Reference \citet{ostrovskii2021finite}.} ( as well as 
\citet{anastasiou2017bounds},
\citet{marteauferey2019beyond})

All these works give \textit{large sample} results -- results that hold only for $n\geq N$ for some constant $N$ -- but none of them  apply to small $n$.

>>>>

\newcommand{\RLP}[1]{\textcolor{red}{RLP:#1}}
\newcommand{\fdk}[1]{\textcolor{Periwinkle}{fdk:#1}}
\newcommand{\TODO}[1]{\textcolor{cyan}{TODO #1}}

% my packages
\usepackage{math_commands}
% some custom math commands
\newtheorem{proposition}{Proposition}
\newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
\newcommand{\cond}{\,\vert\,}
\newcommand{\logpart}{A}
\newcommand{\conj}{\logpart^*}
\newcommand{\bregman}{\cB_\logpart}
\newcommand{\bregmanconj}{\cB_{\logpart^*}}
\newcommand{\natp}{\theta}
\newcommand{\meanp}{\mu}
\newcommand{\decrement}{D}
\newcommand{\linear}{\ell} % linearization of a function
\newcommand{\lr}{\gamma} % learning rate, or step-size
\newcommand{\lin}[1]{\left\langle#1\right\rangle}

\newcommand{\MAPm}{\hat \mu_n}
\newcommand{\MAPt}{\hat \natp_n}

%\hypersetup{draft}

@@

\aistatstitle{Looking for Convergence Rates\\ for the MAP of the Exponential Family}

@@

\begin{abstract}
We consider the problem of upper bounding the expected sub-optimality of the maximum likelihood estimate (MLE), or a conjugate maximum a posteriori (MAP) for the exponential family. 
Surprisingly, we found no solution to this problem in the literature -- e.g. after seeing 5 samples, we do not know how many bits away (in expectation) our model is from the true distribution.
After displaying some properties and special cases of this problem, 
we show it is an application of several optimization algorithms.
Yet it falls out of scope for current analysis, thus highlighting areas for progress.
\end{abstract}

@@

But our understanding of stochastic optimization lacks in other domains. 
A generalization of the above problem, at the intersection of optimization and statistics, 
is the problem of minimizing 
\begin{equation}
	f(\theta) = \expect[x \sim p(x\cond\theta_*)]{-\log p(x\cond\theta_*)},
\end{equation}
When $p(x\cond\theta)$ is an exponential family distribution,
$-\log p(x\cond\theta) = A(\theta) - \lin{t(x), \theta}$ (omitting the reference distribution).
Here, $A$ is the log-partition function, a Legendre function that we assume can be minimized in closed form.
The Gaussian example is the special case $A(\theta) = \frac{1}{2}\norm{\theta}^2$.

There are algorithms for this setting, for example the MLE -- the (online) maximum likelihood after seeing $t$ samples -- or adding regularization, the MAP -- the maximum a posteriori.
As we will prove, the MAP can be interpreted as online mirror descent with decreasing step-sizes.
Despite recent advances in this field, our convergence rates are still lacking ; they are either asymptotic, cover only special cases or require having seen many ($n \gg d$) samples.

@@

We are aware that \emph{a} solution to this problem would be the maximum likelihood estimate for which there is a wealth of asymptotic and large sample results \citep{van2000asymptotic,ostrovskii2021finite}. %TODO complete 
However, none of the theory we found could address for instance the case of a gaussian with unknown mean and variance $\cN(\mu,\sigma^2)$, hence our contributions. 
%TODO kind of redundant "problem - question - problem" paragraphs, not disentangling stats and optim. To improve

\paragraph{Contributions}
\begin{itemize}
	\item We provide insights on different facets of this problem : quadratic approximations ; bias variance decomposition; and a detailed analysis of the gaussian variance estimation problem.
	\item We show the connections between the optimization and statistical estimation view of the problem and discuss where current theories fail to fully capture the problem.
\end{itemize}

@@

Exponential Families are the linear model of probabilities.
Their convexity along with their modeling power make them one of the most widely used statistical models. 
Here we review some of their properties. 
We point the reader towards \citet[Chapter 3]{wainwright2008graphical} for a full review.
The exponential family for data $X\in \cX$ with natural (or primal) parameter $\natp$  is specified by a sufficient statistic $T: \cX \rightarrow \real^d$ and a density with respect to a base measure $\nu(dx)$
\begin{equation}
	 p(X|\natp) = \exp( \natp^\top T(X) - \logpart(\natp)) \; ,
\end{equation}
Here $\logpart$ is the log-partition function -- i.e. the normalization factor
\begin{align}
    \logpart(\natp) = \log \int e^{\natp^\top T(x)} \nu(dx) \; .
\end{align}
When explicit, we will write the random variable $T = T(X)$ for lightness.
This model is defined on the open convex set $\natp\in \Theta$ such that $\logpart(\natp) < \infty$. 
This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}, \nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.

\paragraph{Duality}
The logpartition function $\logpart$ verifies the two following identities
\begin{align}
    \nabla\logpart(\natp) &=  \expect[p(T|\natp)]{T} =: \meanp \\
    \nabla^2 \logpart(\natp) &= \Cov_\natp[T] \geq 0
\end{align}
where $\meanp$ is called the mean (or dual) parameter.
$\meanp$ lives in the open convex set $\cM$ within the convex hull of $T(\cX)$.
%TODO why is it convex ? 
%It is open because it is the reciprocal image of an open set by the continuous mirror map
From now on, we assume that the sufficient statistic $T$ is minimal 
-- e.g. its covariance is definite positive for all $\natp$ -- 
then the log-partition function $\logpart$ is strictly convex and its gradient $\nabla \logpart$ is a bijection between natural parameters $\natp$ and mean parameters $\mu$.
We will write interchangeably $\mu$ or  $\natp$ depending on the context, being aware that both represent the same distribution.
At this point it is useful to introduce the convex conjugate (a.k.a. Fenchel-Legendre transform) of the logpartition function
\begin{align}
	\conj(\mu) = \langle \mu, \natp \rangle - \logpart(\natp) \; .
\end{align}
which matches the common notion of \textit{entropy} in information theory.
The entropy is eating mean parameters, and its gradient is mapping back to natural parameters $\nabla\conj \circ \nabla\logpart(\natp) = \natp$ (cf Fig.~\ref{fig:duality}).

\begin{figure}[ht]
	\centering
	\includegraphics{duality}
	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between natural and mean parameters $\natp, \meanp$. Figure copied from \citet{kunstner2020homeomorphic}. % TODO make an original one.
	}
	\label{fig:duality}
\end{figure}

\paragraph{Bregman Divergences}
A very useful object to keep in mind is the Bregman divergence induced by $\logpart$ between two parameters $\natp$ and $\natp_0$
\begin{align}
    \bregman (\natp ; \natp_0)
    & = \logpart(\natp) - \logpart(\natp_0) 
    - \langle \nabla \logpart(\natp_0)  , \natp - \natp_0 \rangle
\end{align}
with $\nabla \logpart(\natp_0) = \expect[\natp_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\natp_0$. 
It is a non-symmetric measure of distance between parameters.
It is equal to the divergence of its convex conjugate with switched arguments
\begin{align}
	\bregman (\natp ; \natp_0)
    = \bregmanconj ( \meanp_0 ; \meanp) \; .
\end{align}

\paragraph{Conjugate Prior}
One conjugate prior \citep{agarwal2010geometric} for $p(X|\natp)$ is
\begin{align}
    p(\natp) &\propto \exp( - n_0 \bregman(\natp ; \natp_0) )
\end{align}
where $n_0$ and $\natp_0$ are (hyper)parameters of the prior, and $\bregman(\natp ; \natp_0)$ is introduced above.
Intuitively, $n_0$ is a number of fictive points observed from a distribution with parameter $\natp_0$.
The negative log-likelihood of the prior is
\begin{align*}
    -\log p(\natp) = n_0 (\logpart(\natp)  - \natp^\top \meanp_0 ) + \cst
\end{align*}
which is the formula for the exponential family with sufficient statistics $(\natp ,\logpart(\natp))$ and with natural parameter $(n_0 \mu_0, -n_0)$.
Then the joint log-likelihood of $\mD =(X_1,\dots,X_n)$ and $\natp$ is
\begin{multline}
    -\log p(\mD|\natp)p(\natp) 
    = (n_0+n) \logpart (\natp) \\
    - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T_i \right) + \cst
    \label{eq:joint_likelihood}
\end{multline}
where $T_i=T(X_i)$.
This means the posterior given $\mD$ is part of the same family, with natural parameters $(n_0 \mu_0 + \sum_i T_i , -(n_0 + n))$.

\paragraph{Maximum A Posteriori (MAP).}
Minimizing~\eqref{eq:joint_likelihood} over $\natp$ yields the MAP estimate
\begin{align}
    \hat \natp = \argmin_\natp -\log p(\mD|\natp) + n_0 \bregman(\natp ; \natp_0) \; .
\end{align}
If they can be satisfied, the MAP verifies the first oder optimality conditions
\begin{align}
    \nabla \logpart(\hat \natp_\text{MAP}) = \hat \meanp_\text{MAP}
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
\end{align}
When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)
\begin{align}
	\hat \mu_\text{MLE} = \frac{\sum_{i=1}^n T_i}{n}
\end{align}
These rules are known as moment matching.
The MLE and MAP estimates are statistics of the dataset $\mD$. Given a random dataset, we wish to bound their deviation from the optimum $\natp^*$ or $\meanp^*$.

@@

Let $\natp^* := \argmin \expect[X\sim\cD]{-\log p(X|\natp)}$, where $\cD$ is any distribution such that $\E[T] = \meanp^* \in \cM$.
Then the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
\begin{multline}
    \expect[X\sim \cD]{-\log p(X|\natp) + \log p(X|\natp^*) } \\
	= \KL( p(.|\natp^*) ; p(.|\natp)) \; .
\end{multline}
Note that $\cD$ does not have to be part of the exponential family.
For the exponential family, the KL is equal to the Bregman divergence induced by the log-partition function, or alternatively by the entropy, being careful with the order of the arguments 
\begin{align}
\boxed{
	\KL( p(.|\natp^*) ; p(.|\natp))
    = \bregman (\natp ; \natp^*)
    = \bregmanconj ( \meanp^* ; \meanp)
}
\end{align}

@@

Our question is: how does the KL behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get upper-bounds on the following quantities
\begin{align}
	\label{eq:bregmanMLE}
	\expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )} \\
	\label{eq:bregmanMAP}
	\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \mu_0 + \smallsum_i T_i}{n_0+n} \right )} 
\end{align}
where the outer expectation is on the dataset $\mD = \{X_1, \dots, X_n \}$ sampled i.i.d as $X_i\sim \natp^*$?
More explicitly, can we write an upper bound that does not involve an expectation over the dataset -- e.g. where the stochasticity is removed.

\paragraph{Remark.}
This could be seen as a concentration inequality, expressed with a Bregman divergence instead of a norm.
A key difference though, is that the random variable $T(X)$ is connected to the metric $\logpart$. 
Indeed expressions~\eqref{eq:bregmanMLE} or~\eqref{eq:bregmanMAP} can be infinite for another choice of random variable. 
For instance, if we plug in $\conj(\mu)= -\log(\mu)$, which defines a divergence on positive numbers, and $T(X) \sim \cN(0,1)$ which can be negative.

@@

\subsection{Asymptote}
As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
Asymptotic results typically show that natural parameters $\natp$ asymptotically normal ; reaching the Cramer-Rao lower bound \citep[for instance Ch4.2]{van2000asymptotic}.
Here the MAP is more simply expressed with mean parameters $\meanp$.
We get the asymptote by approximating the Bregman divergence using a second order Taylor expansion
\begin{align}
    \bregmanconj(\mu^* ; \mu) 
    &=\half  \norm{\mu^* - \mu}^2_{\nabla^2\conj(\mu^*)}
    + O(\norm{\mu - \mu^*}^3)
\end{align}
where the Mahalanobis norm  $\| x \|_\mM^2 = x^\top \mM x$  is induced by the hessian of the entropy -- the inverse \textit{Fisher information matrix} -- at the optimum, 
\begin{align}
    \mF^*
    :=\nabla^2\conj(\mu^*) 
    = \nabla^2\logpart(\natp^*)^{-1} 
    = \Cov_{\natp^*}[T(X)]^{-1}  \; .
\end{align}
From there, it is possible to show an asymptotic rate for the MLE and the MAP
\begin{align}
\label{eq:MAP_asymptote}
	\E \bregmanconj \left (\E [T(X)] ; \hat \meanp_n^\text{MLE/MAP} \right ) 
	= \frac{d}{2n} + O(n^{- \frac{3}{2}}) \; .
\end{align}
Both MLE and MAP have the same asymptote, as the contribution of the prior $n_0 \meanp_0$ gets negligible compared to the data for large $n$.
This asymptote is also independent of the optimum $\meanp^*$.

\subsection{Euclide's World}
As another reference point, if we make restrictive assumptions on the log-partition function, we can ensure our objective will behave at least as nicely as a quadratic.
\paragraph{If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
\begin{align}
    \bregmanconj(\mu^* ; \mu) 
    &\leq L \norm{\mu^* - \mu} + \norm{\natp} \norm{\mu^* - \mu} \\
    &\leq 2L \norm{\mu^* - \mu}
\end{align}
so $\bregmanconj$ is $2L$-Lipschitz.
\paragraph{If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $L^{-1}$-strongly convex), then
\begin{align}
    \bregmanconj(\mu^* ; \mu) 
    \leq \frac{L}{2} \norm{\mu^* - \mu}^2
\end{align}
so $\bregmanconj$ is upper bounded by a quadratic, and we get the same rate as a Gaussian.

\subsection{Bias-Variance Decomposition}
In both examples (add link), the upper bound takes the form $O(\inv{n}) + O(\frac{\text{bias}}{n^2})$, reminding us of the bias-variance decomposition. In fact, it is possible to write such a decomposition for any Bregman divergence
Let $\tilde \theta_n := \expect{\hat \theta_n}$ be the expectation of the MAP in primal space, and $\tilde \mu_n = \nabla \logpart(\tilde \theta_n )$ be the corresponding mean parameter.
As described by \citet[Theorem 0.1]{pfau2013generalized}, the  expected Bregman decomposes like
\begin{align}
	\expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
	&= {\bregmanconj(\mu^* ; \tilde \mu_n)}
	+ {\expect{\bregmanconj(\tilde \mu_n ; \MAPm)}}
	\label{eq:primal_pivot}
\end{align}

Remark : In this decomposition, the primal expectation $\expect{\hat \theta_n}$ is taking the role of pivot / reference point. An estimator will be unbiased if its primal expectation is non-zero. 
This is not true for the MLE. In fact for the gaussian variance MLE, the bias decreases like ${\bregmanconj(\mu^* ; \tilde \mu_n)} \leq \frac{2}{n(n-2)}$.
A contrario, the MLE is unbiased wrt the dual parameter, but the decomposition there is not clean.

\paragraph{Illustrations.}
We show these decompositions for $\cN(0,\sigma^2)$ in Figure~\ref{fig:variance_decomposition} and  $\cN(\mu, \sigma^2)$ in Figure~\ref{fig:gaussian_decomposition}.
In particular for $\cN(\mu, \sigma^2)$, we illustrate the characters featured in these decompositions $\hat \mu_n^\text{MLE},\hat \mu_n^\text{MAP},\mu_n,\tilde \mu_n, \mu^*$ and $\mu_0$ (and corresponding primal parameters)  in Figure~\ref{fig:bias-variance-numerical}. \RLP{Tune figures.}
